---
title: Getting started with templateflow
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with templateflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: teal
  preset: homage
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',function(){document.body.classList.add('palette-teal');});</script>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Why templateflow?

Neuroimaging analyses almost always need a reference brain template --
an MNI152 T1w image, an atlas parcellation, a surface mesh. The
[TemplateFlow](https://www.templateflow.org/) archive collects these
resources in a single, version-controlled repository with consistent
[BIDS](https://bids-specification.readthedocs.io/) naming.

The **templateflow** R package gives you direct access to this archive
without leaving R. There is no dependency on Python or reticulate --
files are fetched on demand from the public S3 bucket and cached
locally.

```r
library(templateflow)

# one line to get the MNI152 1 mm T1w image
path <- tf_get(template = "MNI152Lin", resolution = 1, suffix = "T1w")
```

## Installation

Install the development version from GitHub:

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("bbuchsbaum/templateflow")
```

For reading NIfTI files directly into R, you will also want one of the
optional neuroimaging packages:

```{r install-optional, eval=FALSE}
install.packages("RNifti")   # lightweight NIfTI reader (recommended)
# or
install.packages("oro.nifti") # alternative NIfTI reader
install.packages("gifti")     # for GIFTI surface files
```

## Discovering templates

Load the package and see which templates are available:

```{r list-templates, eval=FALSE}
library(templateflow)

tf_templates()
#> [1] "Fischer344"        "MNI152Lin"         "MNI152NLin2009cAsym"
#> [4] "MNI152NLin2009cSym" "MNI152NLin6Asym"  "MNI152NLin6Sym"
#> [7] "MNIInfant"         "MNIPediatricAsym"  "NKI"
#> ...
```

The first call initialises a local cache (by default at
`~/.cache/templateflow`) and downloads a lightweight skeleton listing
every file in the archive. Subsequent calls reuse the cache instantly.

> **First run.** If `tf_ls()` or `tf_get()` raises an error mentioning that
> "the cache appears empty," the skeleton failed to populate (often a
> transient network issue). Run `tf_cache_update()` to fetch it explicitly,
> or set `TEMPLATEFLOW_HOME` to point at an existing cache that
> already has the skeleton.

To see what a specific template offers — its citations and the BIDS
entity values you can filter on — use `tf_describe()`:

```{r describe, eval=FALSE}
tf_describe("MNI152Lin")
#> <TemplateFlow description: MNI152Lin>
#>   Name: MNI ICBM 152 linear
#>   References: 1 entry(ies)
#>     - http://nist.mni.mcgill.ca/?p=858
#>
#>   Entities present:
#>     desc       : brain, head, ...
#>     resolution : 1, 2
#>     suffix     : T1w, T2w, mask, probseg, ...
```

## Listing files for a template

Use `tf_ls()` to see what files a template contains. You can filter by
any BIDS entity -- `resolution`, `suffix`, `atlas`, `desc`, `extension`,
and more:

```{r ls-basic, eval=FALSE}
# all files for MNI152Lin
tf_ls(template = "MNI152Lin")

# only the 1 mm T1w image
tf_ls(template = "MNI152Lin", resolution = 1, suffix = "T1w")
```

To see the results as a structured data frame instead of file paths,
pass `as_df = TRUE`:

```{r ls-df, eval=FALSE}
tf_ls(template = "MNI152Lin", as_df = TRUE)
#> # A tibble: 18 x 9
#>    path                template   resolution suffix extension ...
```

You can check which entity names are available for filtering, either
globally or for a specific template:

```{r entities, eval=FALSE}
# all entity names recognised by templateflow
tf_entities()
#>  [1] "atlas"     "cohort"    "density"   "desc"      "extension"
#>  [6] "hemi"      "label"     "resolution" "roi"      "scale"
#> [11] "segmentation" "space"  "suffix"    "template"

# only entities present for this template
tf_entities(template = "MNI152Lin")
#> [1] "desc" "extension" "resolution" "suffix"

# unique values per entity (great for figuring out e.g. which `desc` to use)
tf_entities(template = "MNI152Lin", values = TRUE)$desc
#> [1] "brain" "head"
```

## Fetching files

`tf_get()` works like `tf_ls()` but also downloads any files that are
not yet cached locally. It returns the local file path(s):

```{r get-basic, eval=FALSE}
# download (if needed) and return the local path
path <- tf_get(template = "MNI152Lin", resolution = 1, suffix = "T1w")
path
#> [1] "/Users/you/.cache/templateflow/tpl-MNI152Lin/tpl-MNI152Lin_res-01_T1w.nii.gz"
```

A progress bar appears when multiple files are being downloaded.

By default `tf_get()` raises an error if the query matches no files
(`raise_empty = TRUE`). Pass `raise_empty = FALSE` if you would rather
get back an empty character vector for queries that can legitimately
return zero results.

To get a copy of the files in a directory of your choosing (handy for
shipping templates alongside a project), pass `dest`:

```{r get-dest, eval=FALSE}
tf_get(
  template = "MNI152Lin",
  resolution = 1,
  suffix = "T1w",
  dest = "./templates"
)
#> [1] "./templates/tpl-MNI152Lin/tpl-MNI152Lin_res-01_T1w.nii.gz"
```

The cache remains the canonical store; `dest` receives a snapshot copy
preserving the BIDS sub-path.

### Reading files directly into R

Pass `read = TRUE` to load the file into an R object automatically.
NIfTI files are read with RNifti (or oro.nifti), GIFTI with gifti,
JSON and TSV with their standard readers:

```{r get-read, eval=FALSE}
img <- tf_get(
  template = "MNI152Lin",
  resolution = 1,
  suffix = "T1w",
  read = TRUE
)

img
#> Image array of mode "double" (182 x 218 x 182)
#> - voxel dimensions: 1 x 1 x 1 mm
```

## Template metadata and citations

Every template ships with a `template_description.json` file containing
its name, species, reference space, and citation links. You can access
this with `tf_get_metadata()`:

```{r metadata, eval=FALSE}
meta <- tf_get_metadata(template = "MNI152NLin2009cAsym")
meta$Name
#> [1] "MNI152NLin2009cAsym"

meta$Species
#> [1] "Homo sapiens"
```

For citation information, `tf_get_citations()` returns reference URLs
or, with `bibtex = TRUE`, BibTeX entries you can paste directly into
your bibliography:

```{r citations, eval=FALSE}
tf_get_citations(template = "MNI152NLin2009cAsym")
#> [1] "https://doi.org/10.1016/j.neuroimage.2010.07.033"

tf_get_citations(template = "MNI152NLin2009cAsym", bibtex = TRUE)
```

## Using the global client

You do not need to create or pass a client object explicitly. All
`tf_*()` functions use a default global client behind the scenes. If you
do want to customise settings (e.g. a non-default cache location), create
a client once and set it as the global default:

```{r global-client, eval=FALSE}
client <- TemplateFlowClient(root = "/data/templates")
tf_set_client(client)

# all subsequent calls use /data/templates as the cache
tf_templates()
```

## Cache management

The local cache can be inspected and maintained with a few helpers:

```{r cache, eval=FALSE}
# how big is the cache?
tf_cache_stats()
#> $cache_files
#> [1] 2847
#> $cache_size_human
#> [1] "1.2 GB"

# check for problems (zero-byte stubs, corrupted downloads)
tf_cache_scan()

# re-sync the skeleton from upstream
tf_cache_update()

# rebuild the file index without downloading anything
tf_cache_refresh()

# nuclear option: delete everything and start fresh
tf_cache_wipe()
```

## Configuration

templateflow respects several environment variables:

| Variable | Default | Purpose |
|---|---|---|
| `TEMPLATEFLOW_HOME` | `~/.cache/templateflow` | Cache directory |
| `TEMPLATEFLOW_AUTOUPDATE` | `on` | Check for skeleton updates on first use |
| `TEMPLATEFLOW_TIMEOUT` | `10` (seconds) | Network timeout for downloads |
| `TEMPLATEFLOW_USE_DATALAD` | `off` | Use DataLad backend instead of S3 |

Set these before loading the package, or pass equivalent arguments to
`TemplateFlowClient()`.

## Next steps

- Run `?tf_get` and `?tf_ls` for the full list of entity filters.
- Use `tf_entities()` to discover available BIDS entity names.
- For scripting and shell pipelines, see `?tf_cli` which exposes the
  same functionality as a command-line interface.
- Visit <https://www.templateflow.org/> for documentation on the
  templates themselves.
